---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# NetCoupler

<!-- badges: start -->
[![Join the chat at https://gitter.im/NetCoupler/Lobby](https://badges.gitter.im/NetCoupler/Lobby.svg)](https://gitter.im/NetCoupler/Lobby)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/NetCoupler/NetCoupler.svg?branch=master)](https://travis-ci.org/NetCoupler/NetCoupler)
[![Codecov test coverage](https://codecov.io/gh/NetCoupler/NetCoupler/branch/master/graph/badge.svg)](https://codecov.io/gh/NetCoupler/NetCoupler?branch=master)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/NetCoupler/NetCoupler?branch=master&svg=true)](https://ci.appveyor.com/project/NetCoupler/NetCoupler)
<!-- badges: end -->

The goal of NetCoupler is to estimate causal links between a set of -omic (e.g.
metabolomics, lipidomics) or other high-dimensional metabolic data and either a
disease outcome, an exposure, or both. The NetCoupler-algorithm, formulated by
Clemens Wittenbecher and converted into an R package by Luke Johnston, links
conditional dependency networks with an external outcome or exposure to identify
direct effects between them.

TODO: Add figure demonstrating NetCoupler

See the `vignette("netcoupler.Rmd")` for getting started.

# Installation

So far there is only the development version.

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("NetCoupler/NetCoupler")
```

# Questions? Comments?

- Do you have an informal question, interested in contributing/collaborating, or
want to learn more about NetCoupler? Check out our 
[Gitter Chat Room](https://gitter.im/NetCoupler/Lobby).

# Contributing and Code of Conduct

Checkout the [guidelines](.github/CONTRIBUTING.md) for details on contributing.
Please note that the 'NetCoupler' project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
