---
title: "Introduction to NetCoupler"
author: 
    - "Clemens Wittenbecher"
    - "Luke Johnston"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to NetCoupler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Goals

- Work with pipe `%>%` operator
- Work with tidyselect helpers
- Be autocomplete friendly (e.g. start function names with `nc_`)
- In general, output as tibble, or at least easily modified into a data.frame/tibble
- Arguments are prefixed with `.`
- Internal functions are prefixed with `.`?

## Brief description

The *NetCoupler*-algorithm operates by taking the following steps:

1. The implementation of the PC-algorithm [1] in the pcalg package [@Kalisch2012] is
applied to estimate the skeleton of the underlying directed acyclic graph (DAG)
that generated the complex exposure data (e.g., metabolomics data).
1. For each network variable (e.g., metabolite), the adjacency set (all direct
neighbors) is extracted from this undirected network. Assuming complete coverage
of the measurements, the Markov parents of the network variable are necessarily
a subset of its direct network neighbors. Adjustment for the Markov parents is
sufficient to block all confounding influences by other network-variables
(d-separation). Adjusting for descendants should be avoided because it can
potentially introduce collider bias.
1. A multi-model procedure is applied. The relation of each network variable
with time-to-the event of interest is adjusted for every possible subset of
direct neighbors. Thereby, a confidence set of estimates is generated, which
necessarily contains the valid direct effect estimate.
1. Network variables are classified based on the bounds of this confidence set
of possible direct effects. As default, network variables are classified as
directly affecting the outcome only if the confidence set of possible direct
effects contains exclusively significant and consistent effect estimates.
1. The multi-model procedure is looped with adjusting all models for the
previously identified direct effects (because these are potential confounders
but cannot be colliders) until no further direct effects are identified.

A similar procedure is provided to identify direct influences of external
factors on network variables.

## Workflow

The general workflow for using NetCoupler is to:

- Take the metabolomics data and run the PC-alg on it
- Select a subset of variables or clusters from the PC-alg
- Then run NetCoupler on that subset

## Data pre-processing

Metabolic variables should be log-transformed and standardized on potential
confounders {{this the best way?}} by regressing on age, sex, BMI, and prevalence
of hypertension and using the residual variance in further models. Metabolites were
then scaled (mean of zero and standard deviation of one). {{both regressed and
scaled?}} 

## Directed acyclic graph skeleton estimation

The estimate the underlying directed acyclic graph (DAG) skeleton of the
metabolite network, NetCoupler uses an implementation of the PC-algorithm 
[@Kalisch2012;@Maathuis2010;@Spirtes2001;@Colombo2014]. Because of this, the
assumptions from the PC-algorithm apply to NetCoupler. The output of this
algorithm is a graphical model (`G`) of the underlying network that is used by
NetCoupler.

## Confidence ranges based on multi-model procedures

The estimated DAG skeleton ... {{complete later. Not clear about this right now.
(page 43 of thesis)}}

## Explanation of NetCoupler

Given that 1) an exposure can potentially affect metabolite concentrations
(`exposure -> metabolomics network`) and that 2) metabolites can potentially
affect the development of a disease (`metabolomic network -> outcome`), two initial
algorithms are needed: One for the exposure side (`nc_exposure_network`) and one
for the disease/outcome (`nc_outcome_network`) side. The explanation and
description of NetCoupler are taken heavily from Clemens Wittenbecher's PhD
thesis [@Wittenbecher2017].

### Exposure-side network effect estimation

The exposure-side network retrieves a range of potential direct effects of
exogenous {{word?}} factors on the metabolite network. The general algorithm for
the exposure-side estimation is:

```
input: DAG-skeleton (G)
observations on: network variables (M), exogenous exposures (X), confounders (C)

begin with empty direct effect (DE)
repeat until no further ith M (Mi) is classified as DE
    add new DE to DE
    begin where AMB = M
    repeat until all Mi that are elements of AMB have been selected
        select a variable Mi from AMB
        select all nodes adjacent to Mi in G, adj(G, Mi)
        repeat until no further non-redundant subset (S) can be selected from adj(G, Mi)
            select a S that is an element of adj(G, Mi)
            estimate the predicted Mi as a function of (~) {{confirm}} X
                conditional on S, DE, and C
            add effect estimate PR for X on Mi to the ith confidence set (CS) {{confirm}}
        end
        if the lower bound of CSi > 0 or the upper bound CSi < 0 and
                sign(pe1) = sign(pe2) {{meaning?}} for every pair of estimates in CSi
            classify Mi as affected by X
        else if the lower bound of CSi > 0 or the upper bound of CSi < 0 and 
                0 is an element of CSi or 
                the sign(pe1) is not = sign(pe2) for every pair of estimates in CSi
            classify Mi as ambiguous with respect to X
        else 
            classify Mi as non-affected by X
    end
end

output: CS for effects of X on M based on G
            
classification of every Mi that is an element of M as 
    affected (X -> Mi)
    non-affected (X  Mi)
    ambiguous (X -- Mi)
```

Where `AMB` is {{meaning?}}, `PE` is {{meaning?}}, `adj(G, Mi)` is the set of
nodes adjust to `Mi` in `G` that are congruent ($\equiv$) to direct neighbors of
`Mi` in `G`.

This assumes full source population information. For applications to a sample of
the population, the following decision rules were applied:

1. Consider the exposure-metabolite pair only if associated at a false discovery
rate (FDR) controlled p-value <0.01 based on a model *not adjusted* for the adjacent
metabolites (`S` is empty {{correct?}}, the marginal model).
2. Consider the effect estimates as zero (0) if the p-value is >0.05.

For a given exposure, a metabolite would be classified as non-affected if the
FDR-adjusted p-value was non-significant in the marginal model. If there was
a significant FDR-controlled p-value in the marginal model and all estimates in
the confidence set (`CS` {{? correct?}}) were significant and consistent, then
the metabolite would be classified as affected by the exposure. Other metabolites
would be classified as ambiguous with respect to the exposure.

{{TODO: reduce reliance on p-values or use another metric? Or do some bootstrapping/CV?}}

The outer loop includes identifying directly affected metabolites into the fixed
model part. Then the procedure was repeated on the still ambiguous metabolites 
to check whether further unambiguous classification was possible based on the
additional information. In the applied version of NetCoupler, this was limited
to metabolites in the same connected components, which are defined as a group of
two or more directly linked metabolites that were all associated with the
exposure in the marginal model.

The rationale was that indirect effects could only be mediated by metabolites
that were themselves (directly or indirectly) affected by the exposure. These
metabolites were expected to be marginally associated with the exposure.
In theory, there could be possible scenarios where the incidental cancellation
of several direct and indirect effects concealed exposure-dependency in the 
marginal model [@Pearl2009]

Still, high abundance of incidental cancellations are considered unlikely based
on observed correlation structures and the chance to unambiguously resolve such
complicated scenarios in the applied modelling approach is low {{proof of
claim?}}. Therefore, the pragmatic decision was taken to adjust remaining
ambiguous metabolites only for directly affected metabolites identified within
the same connected component.

{{TODO: rearrange this}}

Network adjustments were used to resolve indirect effects mediated by another
network variable (e.g. `exposure -> mediating metabolite -> metabolite`).

### Outcome-side network effect estimation

The outcome-side network retrieves direct effects of network variables on later
occurring events.

There are similarities to the exposure-side algorithm for the outcome-side
algorithm, except that the directionality is reversed (`metabolite -> outcome`).
This change in directionality also has implications on the decision rules.
There may also be potential network-variable confounders and introduction
of spurious associations (`metabolite <- confounding metabolite -> outcome`).
The general algorithm for the outcome-side estimation is:

```
input: DAG-skeleton G (Graphical model)
observations on: network variables (M), outcome (O), confounders (C)

begin with empty direct effect (DE)
repeat until no further ith M (Mi) is classified as DE
    add new DE to DE
    begin where AMB = M
    repeat until all Mi have been selected
        select a variable Mi from M {{not AMB?}}
        select all nodes adjacent to Mi in G, adj(G, Mi)
        repeat until no further non-redundant subset (S) can be selected from adj(G, Mi)
            select a S from adj(G, Mi)
            estimate outcome as a function of Mi conditional on S, C
            add effect estimate PE for Mi on O to confidence set (CSi)
        end
        if the lower bound of CSi > 0 or upper bound of CSi < 0 and
                sign(pe1) = sign(pe2) for every pair of estimates in CSi
            classify Mi as affecting risk of O
        else if lower bound of CSi > 0 or upper bound of CSi < 0 and
                0 is in CSi or 
                sign(pe1) = sign(pe2) for any pair of estimates in CSi
            classify Mi as ambiguous with respect to risk of O
        else
            classify Mi as non-affecting risk to O
    end
end

output: CS for effects of M on O based on G

classification of every Mi from M as 
    effector on O (Mi -> O)
    non-effector on O (Mi  O)
    ambiguous (Mi -- O)
```

Where `AMB` is {{meaning?}}, `PE` is {{meaning?}}, `adj(G, Mi)` is the set of
nodes adjust to `Mi` in `G` that are congruent ($\equiv$) to direct neighbors of
`Mi` in `G`.

An indirect effect, unlike confounding, is not biased. Decision rules were adjusted
to avoid bias:

1. Consider a metabolite-outcome pair only if significantly associated at a
FDR-adjusted p-value <0.1 based on a model adjusted for all adjacent metabolites
(`S = adj(Mi)`)
1. Consider effect estimates as 0 if p-value is >0.05

Metabolites are classified as affecting the outcome if the FDR-adjusted p-value
was significant in the model adjusted for the full adjacency set and all estimates
in the confidence set are significant and consistent. Metabolites are considered
not directly affecting the outcome if they were not significantly associated in
the model adjusted for the full adjacency set. All others were considered
ambiguous.

### Step-wise iteration for full network estimation

1. The DAG skeleton of the metabolite network is estimated
2. Initial, tentative links are formed with the exposure-side of the network.
3. Direct effects are estimated between exposure and network.
4. Repeat identification of links, classifying as unaffected or ambiguous and 
deleting indirect links.
5. Initial, tentative links are formed with the outcome-side of the network.
6. Direct effects are estimated between the network and the outcome.
7. Repeat link identification, deleting indirect links and classifying others as
unaffected or ambiguous.
8. Combine the exposure-side and outcome-side network estimates into a joint 
graphical model.


<!-- TODO: Put this into a bib file. -->

1. *Spirtes P, Glymour C. An Algorithm for Fast Recovery of Sparse Causal Graphs.
Social Science Computer Review 1991;9:62-72.*

<!-- thesis https://publishup.uni-potsdam.de/frontdoor/index/index/searchtype/collection/id/17105/start/1/rows/10/facetNumber_author_facet/all/doctypefq/doctoralthesis/docId/40459 -->

2 *Markus Kalisch, Martin Maechler, Diego Colombo, Marloes H. Maathuis, Peter
Buehlmann (2012). Causal Inference Using Graphical Models with the R Package
pcalg. Journal of Statistical Software, 47(11), 1-26. URL
http://www.jstatsoft.org/v47/i11/.*

## Simple example of use

```{r load-packages, message=FALSE, warning=FALSE}
library(NetCoupler)
library(dplyr)
```

### Estimating the metabolic network

For estimating the network, it's (basically) required to standardize
the metabolic variables before inputting into `nc_estimate_network()`.
If you intend to also adjust for potential confounders when estimating
the exposure or outcome side connections,
you can include the potential impact these confounders may have on 
the network by regressing the confounders on the metabolic variables.
Then the residuals can be extracted and used when constructing the network.
You do this with the `nc_standardize()` function. 
This function also log-transforms and scales 
(mean-center and z-score normalize) the values of the metabolic variables.
We do this because the network estimation algorithm can sometimes be finicky
about differences in variable numerical scale (mean of 1 vs mean of 1000).

```{r metabolic-residuals}
std_metabolic_data <- simulated_data %>% 
    nc_standardize(starts_with("metabolite"), 
                   .regressed_on = "age") %>% 
    select(starts_with("metabolite"))
```

After that, you can estimate the network.

```{r create-network}
# Make partial independence network from metabolite data
metabolite_network <- std_metabolic_data %>% 
    nc_estimate_network()
```

To see what the network looks like,
use the function `nc_plot_network()`.

```{r visualize-metabolic-network, fig.width=5.6, fig.height=4.5}
std_metabolic_data %>%
    nc_plot_network(metabolite_network)
```

The plot is a bit crowded, but it provides a base to start tidying up from.

### Estimating exposure and outcome-side connections

For the exposure and outcome side, 
you should standardize the metabolic variables, 
but this time, we don't regress on the confounders 
since they will be included in the models.

```{r standardize-data}
standardized_data <- simulated_data %>% 
    nc_standardize(starts_with("metabolite"))
```

Then estimate the outcome or exposure:

```{r example-use, cache=TRUE}
outcome_estimates <- standardized_data %>%
    nc_outcome_estimates(
        .edge_tbl = as_edge_tbl(metabolite_network),
        .outcome = "case_status",
        .model_function = glm,
        .model_arg_list = list(family = binomial(link = "logit")),
        .exponentiate = TRUE
    )

outcome_estimates

exposure_estimates <- standardized_data %>%
    nc_exposure_estimates(
        .edge_tbl = as_edge_tbl(metabolite_network),
        .exposure = "exposure",
        .adjustment_vars = "age",
        .model_function = lm
    )

exposure_estimates
```

### Classifying direct effects on exposure or outcome side

You can classify direct effects by using `nc_classify_effects()`,
which works on both the exposure side (`exposure -> metabolite`) 
and the outcome side (`metabolite -> outcome`).
The function applies a simple algorithm, described in more
detail in `vignette("description")`.
For the exposure side, 
the function identifies whether a link between the exposure 
and an index node (one metabolic variable in the network) exists,
regardless of conditioning on potential confounders 
and on neighbouring nodes (other metabolic variables linked to the index variable).
Depending on how consistent and strong the link is,
the effect is classified as "direct", "ambiguous", or "none".

In the example below, we specifically generated the simulated data so that
the exposure associated with the metabolites 1, 8, and 12.
And as we can see, those links have been correctly identified.

```{r classify-exposure-effects}
exposure_direct_effects <- exposure_estimates %>% 
    nc_classify_effects()
exposure_direct_effects
exposure_direct_effects %>% 
    dplyr::filter(direct_effect == "direct")
```

You can also classify direct effects with an outcome using the same function:

```{r classify-outcome-effects}
outcome_direct_effects <- outcome_estimates %>% 
    nc_classify_effects()
outcome_direct_effects
outcome_direct_effects %>% 
    dplyr::filter(direct_effect == "direct")
```

### Plotting 

To visualize the results of the linked network graph and the effect
classification, there are two functions to show the exposure and the outcome
plots. In general, these plot functions are currently mostly for exploratory
purposes and are too "busy" and crowded to be meaningful for presentation or
publication. However, these are good starting points for making prettier,
more legible plots.

```{r plot-outcome-estimation-networks, fig.width=9, fig.height=8}
nc_plot_outcome_estimation(
    standardized_data,
    metabolite_network,
    outcome_direct_effects
)
```

```{r plot-exposure-estimation-networks, fig.width=9, fig.height=8}
nc_plot_exposure_estimation(
    standardized_data,
    metabolite_network,
    exposure_direct_effects
)
```



```{r, eval=FALSE}
standardized_data <- simulated_data %>% 
    nc_standardize(starts_with("metabolite"))

metabolite_network <- simulated_data %>% 
    nc_standardize(starts_with("metabolite"), 
                   .regressed_on = "age") %>% 
    nc_estimate_network(starts_with("metabolite"))
edge_table <- as_edge_tbl(metabolite_network)
# future::plan(future::multisession)
exposure_estimates <- standardized_data %>%
    nc_exposure_estimates(
        .edge_tbl = edge_table,
        .exposure = "exposure",
        .adjustment_vars = "age",
        .model_function = lm
    )
exposure_direct_effects <- exposure_estimates %>% 
    nc_classify_effects()
de_vars <- exposure_direct_effects %>% 
    filter(direct_effect == "direct") %>% 
    pull(index_node) %>% 
    unique()

edge_table_2 <- as_edge_tbl(metabolite_network) %>% 
    filter(!source_node %in% de_vars)
exposure_estimates_with_de <- standardized_data %>%
    nc_exposure_estimates(
        .edge_tbl = edge_table_2,
        .exposure = "exposure",
        .adjustment_vars = "age",
        .direct_effect_vars = de_vars,
        .model_function = lm
    )
exposure_direct_effects_2 <- exposure_estimates_with_de %>% 
    nc_classify_effects()

View(exposure_estimates)
View(exposure_estimates_with_de)

View(exposure_direct_effects)
View(exposure_direct_effects_2)

exposure_direct_effects_2 %>% 
    View()
    pull(index_node) %>% 
    unique()
    
de_vars_2 <- exposure_direct_effects_2 %>% 
    filter(direct_effect == "direct") %>% 
    pull(index_node) %>% 
    unique()

edge_table_3 <- as_edge_tbl(metabolite_network) %>% 
    filter(!source_node %in% c(de_vars, de_vars_2))
           
exposure_estimates_with_de_2 <- standardized_data %>%
    nc_exposure_estimates(
        .edge_tbl = edge_table_3,
        .exposure = "exposure",
        .adjustment_vars = c("age", de_vars, de_vars_2),
        .model_function = lm
    )
exposure_direct_effects_3 <- exposure_estimates_with_de_2 %>% 
    nc_classify_effects()

```

