---
title: "Getting started with NetCoupler"
author: 
    - "Clemens Wittenbecher"
    - "Luke Johnston"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with NetCoupler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

![](https://img.shields.io/badge/document%20status-rough%20draft-red?style=flat-square)

## Brief description

The goal of NetCoupler is to estimate causal links between a set of -omic (e.g.
metabolomics, lipidomics) or other high-dimensional metabolic data and either a
disease outcome, an exposure, or both. The NetCoupler-algorithm, formulated by
Clemens Wittenbecher and converted into an R package by Luke Johnston, links
conditional dependency networks with an external outcome or exposure to identify
direct effects between them.

The NetCoupler's input is multi-layer information from prospective studies,
including interdependent variables that constitute the central network of
interest (e.g., metabolomics data), time-to-disease incidence, and optionally
information on factors that influence the network (such as lifestyle variables,
or genetic determinants).

TODO: Add figure demonstrating NetCoupler

Results can be graphically displayed as a joint network model. For example, to a
data-driven metabolomics network links can be added that reflect
network-independent associations of metabolites with disease risk and lifestyle
habits (or genetics) with these disease-related metabolites.


## Workflow

The general workflow for using NetCoupler is to:

1. Standardize variables that will form the network ("network variables") with
`nc_standardize()`.
1. Estimate the network linkages for the network variables with
`nc_create_network()`.
1. If interested in links to an *outcome*: Using the estimated network, extract
the network variables (both "index" and "neighbour" variable list) and identify
direct and ambiguous effects from the network to the outcome variable with
`nc_outcome_estimates()`.
1. If interested in links to an *exposure*: Just like with the *outcome*-side,
except identify direct and ambiguous effects from the exposure variable to the
network `nc_exposure_estimates()`.


1. Using the estimated network, extract the network variables (both "index" and
"neighbour" variable list) and identify direct and ambiguous effects with the
outcome variable with `nc_outcome_estimates()`. 
- Clusters of the variables will be selected based on the PC-algorithm
Select a subset of variables or clusters from the PC-alg
- Then run NetCoupler on that subset

## Directed acyclic graph skeleton estimation

The estimate the underlying directed acyclic graph (DAG) skeleton of the
metabolite network, NetCoupler uses an implementation of the PC-algorithm 
[@Kalisch2012;@Maathuis2010;@Spirtes2001;@Colombo2014]. Because of this, the
assumptions from the PC-algorithm apply to NetCoupler. The output of this
algorithm is a graphical model (`G`) of the underlying network that is used by
NetCoupler.

## Simple example of use

```{r load-packages, message=FALSE, warning=FALSE}
library(NetCoupler)
library(dplyr)
```

### Estimating the metabolic network

For estimating the network, it's (basically) required to standardize
the metabolic variables before inputting into `nc_create_network()`.
If you intend to also adjust for potential confounders when estimating
the exposure or outcome side connections,
you can include the potential impact these confounders may have on 
the network by regressing the confounders on the metabolic variables.
Then the residuals can be extracted and used when constructing the network.
You do this with the `nc_standardize()` function. 
This function also log-transforms and scales 
(mean-center and z-score normalize) the values of the metabolic variables.
We do this because the network estimation algorithm can sometimes be finicky
about differences in variable numerical scale (mean of 1 vs mean of 1000).

```{r metabolic-residuals}
std_metabolic_data <- simulated_data %>% 
    nc_standardize(starts_with("metabolite"), 
                   .regressed_on = "age") %>% 
    select(starts_with("metabolite"))
```

After that, you can estimate the network.

```{r create-network}
# Make partial independence network from metabolite data
metabolite_network <- std_metabolic_data %>% 
    nc_create_network()
```

To see what the network looks like,
use the function `nc_plot_network()`.

```{r visualize-metabolic-network, fig.width=5.6, fig.height=4.5}
std_metabolic_data %>%
    nc_plot_network(metabolite_network)
```

The plot is a bit crowded, but it provides a base to start tidying up from.

### Estimating exposure and outcome-side connections

For the exposure and outcome side, 
you should standardize the metabolic variables, 
but this time, we don't regress on the confounders 
since they will be included in the models.

```{r standardize-data}
standardized_data <- simulated_data %>% 
    nc_standardize(starts_with("metabolite"))
```

Then estimate the outcome or exposure:

```{r example-use, cache=TRUE}
outcome_estimates <- standardized_data %>%
    nc_outcome_estimates(
        .graph = metabolite_network,
        .outcome = "case_status",
        .model_function = glm,
        .model_arg_list = list(family = binomial(link = "logit")),
        .exponentiate = TRUE
    )

outcome_estimates

exposure_estimates <- standardized_data %>%
    nc_exposure_estimates(
        .graph = metabolite_network,
        .exposure = "exposure",
        .adjustment_vars = "age",
        .model_function = lm
    )

exposure_estimates
```

### Classifying direct effects on exposure or outcome side

You can classify direct effects by using `nc_classify_effects()`,
which works on both the exposure side (`exposure -> metabolite`) 
and the outcome side (`metabolite -> outcome`).
The function applies a simple algorithm, described in more
detail in `vignette("description")`.
For the exposure side, 
the function identifies whether a link between the exposure 
and an index node (one metabolic variable in the network) exists,
regardless of conditioning on potential confounders 
and on neighbouring nodes (other metabolic variables linked to the index variable).
Depending on how consistent and strong the link is,
the effect is classified as "direct", "ambiguous", or "none".

In the example below, we specifically generated the simulated data so that
the exposure associated with the metabolites 1, 8, and 12.
And as we can see, those links have been correctly identified.

```{r classify-exposure-effects}
exposure_direct_effects <- exposure_estimates %>% 
    nc_classify_effects()
exposure_direct_effects
exposure_direct_effects %>% 
    dplyr::filter(direct_effect == "direct")
```

You can also classify direct effects with an outcome using the same function:

```{r classify-outcome-effects}
outcome_direct_effects <- outcome_estimates %>% 
    nc_classify_effects()
outcome_direct_effects
outcome_direct_effects %>% 
    dplyr::filter(direct_effect == "direct")
```

### Plotting 

To visualize the results of the linked network graph and the effect
classification, there are two functions to show the exposure and the outcome
plots. In general, these plot functions are currently mostly for exploratory
purposes and are too "busy" and crowded to be meaningful for presentation or
publication. However, these are good starting points for making prettier,
more legible plots.

```{r plot-outcome-estimation-networks, fig.width=9, fig.height=8}
nc_plot_outcome_estimation(
    standardized_data,
    metabolite_network,
    outcome_direct_effects
)
```

```{r plot-exposure-estimation-networks, fig.width=9, fig.height=8}
nc_plot_exposure_estimation(
    standardized_data,
    metabolite_network,
    exposure_direct_effects
)
```

