---
title: "Introduction to NetCoupler"
author: 
    - "Clemens Wittenbecher"
    - "Luke Johnston"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to NetCoupler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Brief description

The *NetCoupler*-algorithm operates by taking the following steps:

1. The implementation of the PC-algorithm [1] in the pcalg package [2] is
applied to estimate the skeleton of the underlying directed acyclic graph (DAG)
that generated the complex exposure data (e.g., metabolomics data).
1. For each network variable (e.g., metabolite), the adjacency set (all direct
neighbors) is extracted from this undirected network. Assuming complete coverage
of the measurements, the Markov parents of the network variable are necessarily
a subset of its direct network neighbors. Adjustment for the Markov parents is
sufficient to block all confounding influences by other network-variables
(d-separation). Adjusting for descendants should be avoided because it can
potentially introduce collider bias.
1. A multi-model procedure is applied. The relation of each network variable
with time-to-the event of interest is adjusted for every possible subset of
direct neighbors. Thereby, a confidence set of estimates is generated, which
necessarily contains the valid direct effect estimate.
1. Network variables are classified based on the bounds of this confidence set
of possible direct effects. As default, network variables are classified as
directly affecting the outcome only if the confidence set of possible direct
effects contains exclusively significant and consistent effect estimates.
1. The multi-model procedure is looped with adjusting all models for the
previously identified direct effects (because these are potential confounders
but cannot be colliders) until no further direct effects are identified.

A similar procedure is provided to identify direct influences of external
factors on network variables.

1 *Spirtes P, Glymour C. An Algorithm for Fast Recovery of Sparse Causal Graphs.
Social Science Computer Review 1991;9:62-72.*
2 *Markus Kalisch, Martin Maechler, Diego Colombo, Marloes H. Maathuis, Peter
Buehlmann (2012). Causal Inference Using Graphical Models with the R Package
pcalg. Journal of Statistical Software, 47(11), 1-26. URL
http://www.jstatsoft.org/v47/i11/.*

# Code usage (from original scripts)

Need to go through and organize this code.

```{r}

# metabolite data (e.g. phosphatidylcholines) data for subcohort:

met_data_SC <- readRDS("STR_GPL_SC.rds")
Exp_data_SC <- readRDS("EXP_SC.rds") # phenotype data for 2092 patients, including diet information, lifestyle, etc.


met_data_SC_rename <- rename.met(dat = met_data_SC)$data_renamed # rename metabolites with short names
met_mapping <- rename.met(dat = met_data_SC)$names_mapping # mapping information between old and new metabolite names

met_estimates_norename_005 <- est.pcor.skel.DAG.adj(dat = met_data_SC, alpha_val = 0.05)
met_skel_norename_005 <- met_estimates_norename_005$skel_est # estimate DAG skeleton for non-renamed matrix
met_DAG_norename_005 <- met_estimates_norename_005$DAG_est # estimate DAG for non-renamed matrix
met_adj_mat_norename_005 <- met_estimates_norename_005$adj_matrix # estimate adjacency matrix for non-renamed matrix

# estimate direct effects of predefined exposure on each network-variable, causal models that agree with the input-network: models are adjusted for all possible combinations of direct neighbors (==variables in adjacency set) -> Output is multiset of possible effects:
net_coupler_in_PC_005 <- net.coupler.in(graph_skel = met_skel_rename_005, dat = met_data_SC_rename, dat_compl = met_data_SC_rename, exp_dat = Exp_data_SC, DE = NULL, glmulti_method = "h")

# 1.) return results (e.g. p-values, etc.) for whole-grain bread:
sum_netin_WGB_005 <- getExp.coef(object = net_coupler_in_PC_005, outcome = colnames(met_data_SC_rename), exposure = "WGBperMJ")

# get original metabolite names back:
sum_netin_WGB_005 <- merge(sum_netin_WGB_005, as.data.frame(met_mapping), by = "Outcome")
sum_stat_netin_WGB_005 <- mult.stat(sum_netin = sum_netin_WGB_005, MinMod = MinMod_exp, adjust_method = "fdr", round_number = 1) # calculate summary statistics and determine direct and ambiguous effects

netin_WGB_amb_005 <- sum_stat_netin_WGB_005$amb # summary statistics for metabolites on which WGB has ambiguous effect
netin_WGB_direct_005 <- sum_stat_netin_WGB_005$direct # summary statistics for metabolites on which WGB has direct effect
netin_WGB_sum_005 <- sum_stat_netin_WGB_005$sum_netin_sum_adj_FV # summary statistics for all metabolites including round-number information
# 2.) return results (e.g. p-values, etc.) for redmeat:
sum_netin_redmeat_005 <- getExp.coef(object = net_coupler_in_PC_005, outcome = colnames(met_data_SC_rename), exposure = "TMperMJ")
# get original metabolite names back:
sum_netin_redmeat_005 <- merge(sum_netin_redmeat_005, as.data.frame(met_mapping), by = "Outcome")
sum_stat_netin_redmeat_005 <- mult.stat(sum_netin = sum_netin_redmeat_005, MinMod = MinMod_exp, adjust_method = "fdr", round_number = 1) # calculate summary statistics and determine direct and ambiguous effects
netin_redmeat_amb_005 <- sum_stat_netin_redmeat_005$amb # summary statistics for metabolites on which redmeat has ambiguous effect
netin_redmeat_direct_005 <- sum_stat_netin_redmeat_005$direct # summary statistics for metabolites on which redmeat has direct effect
netin_redmeat_sum_005 <- sum_stat_netin_redmeat_005$sum_netin_sum_adj_FV # summary statistics for meatbolites including round-number information

# there are direct effects of redmeat on metabolites -> extract connected components:

# extract connected components for direct effects for redmeat:
con_comp_redmeat_005 <- get.con.comp(exposure_names = c("Redmeat"), exposure_list = list(Redmeat = netin_redmeat_sum_005), adjM_norename = met_adj_mat_norename_005, met_group = "PC")

con_comp_redmeat_0001 <- get.con.comp(exposure_names = c("Redmeat"), exposure_list = list(Redmeat = netin_redmeat_sum_0001), adjM_norename = met_adj_mat_norename_0001, met_group = "PC")


# repeat net.coupler.in with new set of fixed variables ("always"-set consists now of all exposures, all covariates, and all connected components direct effects):
net_coupler_in_PC_redmeat_CC2_005 <- net.coupler.in(graph_skel = met_skel_rename_005, dat = met_data_SC_rename_AMB1_redmeat_CC2_005, dat_compl = met_data_SC_rename, exp_dat = Exp_data_SC_DE1_redmeat_CC2_005, DE = DE1_redmeat_CC2_005_rename, glmulti_method = "h")
# return results (e.g. p-values, etc.) for redmeat for second net.coupler.in round for CC2:
sum_netin_redmeat_CC2_005 <- getExp.coef(object = net_coupler_in_PC_redmeat_CC2_005, outcome = colnames(met_data_SC_rename_AMB1_redmeat_CC2_005), exposure = "TMperMJ")

sum_stat_netin_redmeat_CC2_005 <- mult.stat(sum_netin = sum_netin_redmeat_CC2_005, MinMod = MinMod_exp_DE1_redmeat_CC2_005, adjust_method = "fdr", round_number = 2) # calculate summary statistics and determine direct and ambiguous effects


# repeat net.coupler.in with new set of fixed variables ("always"-set consists now of all exposures, all covariates, and all connected components direct effects):
net_coupler_in_PC_redmeat_CC3_005 <- net.coupler.in(graph_skel = met_skel_rename_005, dat = met_data_SC_rename_AMB1_redmeat_CC3_005, dat_compl = met_data_SC_rename, exp_dat = Exp_data_SC_DE1_redmeat_CC3_005, DE = DE1_redmeat_CC3_005_rename, glmulti_method = "h")


# return results (e.g. p-values, etc.) for redmeat for second net.coupler.in round for CC3:
sum_netin_redmeat_CC3_005 <- getExp.coef(object = net_coupler_in_PC_redmeat_CC3_005, outcome = colnames(met_data_SC_rename_AMB1_redmeat_CC3_005), exposure = "TMperMJ")

sum_stat_netin_redmeat_CC3_005 <- mult.stat(sum_netin = sum_netin_redmeat_CC3_005, MinMod = MinMod_exp_DE1_redmeat_CC3_005, adjust_method = "fdr", round_number = 2) # calculate summary statistics and determine direct and ambiguous effects

# 3.) return results (e.g. p-values, etc.) for coffee:
sum_netin_coffee_005 <- getExp.coef(object = net_coupler_in_PC_005, outcome = colnames(met_data_SC_rename), exposure = "CofCup")
sum_stat_netin_coffee_005 <- mult.stat(sum_netin = sum_netin_coffee_005, MinMod = MinMod_exp, adjust_method = "fdr", round_number = 1) # calculate summary statistics and determine direct and ambiguous effects

# there are direct effects of coffee on metabolites -> extract connected components:

# extract connected components for direct effects for coffee:
con_comp_coffee_005 <- get.con.comp(exposure_names = c("Coffee"), exposure_list = list(Coffee = netin_coffee_sum_005), adjM_norename = met_adj_mat_norename_005, met_group = "PC")

# visualisation of skeletons and DAGs?

# 1.) return results (e.g. p-values, etc.) for whole-grain bread:

net_coupler_in_WGB_005 <- amb.met.loop.CC(exp_dat = Exp_data_SC, graph_skel = met_skel_rename_005, dat = met_data_SC_rename, dat_compl = met_data_SC_rename, DE = NULL, glmulti_method = "h", exposure = "WGBperMJ", met_map = met_mapping, adjust_method = "fdr", round_number = 1, adjM_norename = met_adj_mat_norename_005, met_group = "PC")
# no direct effect of WGB on metabolites -> ambiguous effects still classified as ambiguous

net_coupler_in_WGB_0001 <- amb.met.loop.CC(exp_dat = Exp_data_SC, graph_skel = met_skel_rename_0001, dat = met_data_SC_rename, dat_compl = met_data_SC_rename, DE = NULL, glmulti_method = "h", exposure = "WGBperMJ", met_map = met_mapping, adjust_method = "fdr", round_number = 1, adjM_norename = met_adj_mat_norename_0001, met_group = "PC")
# 2.) return results (e.g. p-values, etc.) for redmeat:

net_coupler_in_redmeat_005 <- amb.met.loop.CC(exp_dat = Exp_data_SC, graph_skel = met_skel_rename_005, dat = met_data_SC_rename, dat_compl = met_data_SC_rename, DE = NULL, glmulti_method = "h", exposure = "TMperMJ", met_map = met_mapping, adjust_method = "fdr", round_number = 1, adjM_norename = met_adj_mat_norename_005, met_group = "PC")

met_data_SC_rename <- rename.met(dat = met_data_SC)$data_renamed # rename metabolites with short names
met_mapping <- rename.met(dat = met_data_SC)$names_mapping # mapping information between old and new metabolite names

met_estimates_norename <- est.pcor.skel.DAG.adj(dat = met_data_SC)
met_skel_norename <- met_estimates_norename$skel_est # estimate DAG skeleton for non-renamed matrix
met_DAG_norename <- met_estimates_norename$DAG_est # estimate DAG for non-renamed matrix
met_adj_mat_norename <- met_estimates_norename$adj_matrix # estimate adjacency matrix for non-renamed matrix

met_estimates_rename <- est.pcor.skel.DAG.adj(dat = met_data_SC_rename)
met_skel_rename <- met_estimates_rename$skel_est # estimate DAG skeleton for renamed matrix
met_DAG_rename <- met_estimates_rename$DAG_est # estimate DAG for renamed matrix
met_adj_mat_rename <- met_estimates_rename$adj_matrix # estimate adjacency matrix for renamed matrix

# visualisation of skeletons and DAGs?

# longer analysis version#

MinMod_exp <- c(paste(colnames(Exp_data_SC), collapse = ", ")) # record covariates for minimal model without any adjacency set

# estimate direct effects of predefined exposure on each network-variable, causal models that agree with the input-network: models are adjusted for all possible combinations of direct neighbors (==variables in adjacency set) -> Output is multiset of possible effects:
net_coupler_in_PC <- net.coupler.in(graph_skel = met_skel_rename, dat = met_data_SC_rename, dat_compl = met_data_SC_rename, exp_dat = Exp_data_SC, DE = NA)

# 1.) return results (e.g. p-values, etc.) for whole-grain bread:
sum_netin_WGB <- getExp.coef(object = net_coupler_in_PC, outcome = colnames(met_data_SC_rename), exposure = "WGBperMJ")

# get original metabolite names back:
sum_netin_WGB <- merge(sum_netin_WGB, as.data.frame(met_mapping), by = "Outcome")

sum_stat_netin_WGB <- mult.stat(sum_netin = sum_netin_WGB, MinMod = MinMod_exp, adjust_method = "fdr", round_number = 1) # calculate summary statistics and determine direct and ambiguous effects


# extract connected components for direct effects for redmeat:
con_comp_redmeat <- get.con.comp(exposure_names = c("Redmeat"), exposure_list = list(Redmeat = netin_redmeat_sum), adjM_norename = met_adj_mat_norename, met_group = "PC")

# no direct effect of WGB on metabolites -> ambiguous effects still classified as ambiguous

# 2.) return results (e.g. p-values, etc.) for redmeat:

net_coupler_in_redmeat <- amb.met.loop.CC(exp_dat = Exp_data_SC, graph_skel = met_skel_rename, dat = met_data_SC_rename, dat_compl = met_data_SC_rename, DE = NULL, exposure = "TMperMJ", met_map = met_mapping, adjust_method = "fdr", round_number = 1, adjM_norename = met_adj_mat_norename, met_group = "PC")

met_data_SC_rename <- rename.met(dat = met_data_SC)$data_renamed # rename metabolites with short names
met_mapping_SC <- rename.met(dat = met_data_SC)$names_mapping # mapping information between old and new metabolite names

met_data_rename <- rename.met(dat = met_data)$data_renamed # rename metabolites with short names
met_mapping <- rename.met(dat = met_data)$names_mapping # mapping information between old and new metabolite names

# skeleton estimation for metabolite matrix subcohort:
met_SC_estimates_norename <- est.pcor.skel.DAG.adj(dat = met_data_SC)
met_SC_skel_norename <- met_SC_estimates_norename$skel_est # estimate DAG skeleton for non-renamed matrix
met_SC_DAG_norename <- met_SC_estimates_norename$DAG_est # estimate DAG for non-renamed matrix
met_SC_adj_mat_norename <- met_SC_estimates_norename$adj_matrix # estimate adjacency matrix for non-renamed matrix

met_SC_estimates_rename <- est.pcor.skel.DAG.adj(dat = met_data_SC_rename)
met_SC_skel_rename <- met_SC_estimates_rename$skel_est # estimate DAG skeleton for renamed matrix
met_SC_DAG_rename <- met_SC_estimates_rename$DAG_est # estimate DAG for renamed matrix
met_SC_adj_mat_rename <- met_SC_estimates_rename$adj_matrix # estimate adjacency matrix for renamed matrix

# skeleton estimation for complete cohort metabolite matrix:
met_estimates_norename <- est.pcor.skel.DAG.adj(dat = met_data)
met_skel_norename <- met_estimates_norename$skel_est # estimate DAG skeleton for non-renamed matrix
met_DAG_norename <- met_estimates_norename$DAG_est # estimate DAG for non-renamed matrix
met_adj_mat_norename <- met_estimates_norename$adj_matrix # estimate adjacency matrix for non-renamed matrix

met_estimates_rename <- est.pcor.skel.DAG.adj(dat = met_data_rename)
met_skel_rename <- met_estimates_rename$skel_est # estimate DAG skeleton for renamed matrix
met_DAG_rename <- met_estimates_rename$DAG_est # estimate DAG for renamed matrix
met_adj_mat_rename <- met_estimates_rename$adj_matrix # estimate adjacency matrix for renamed matrix

# create "survival" object:
t2d_surv <- Surv(T2D_data$sta_time, T2D_data$sto_time, T2D_data$fall)


# initial net.coupler.out run:
net_coupler_out_1 <- net.coupler.out(graph_skel = met_SC_skel_rename, dat = met_data_rename, dat_compl = met_data_rename, exp_dat = Exp_data, DE = NULL, survival_obj = t2d_surv, always_set = always_set)

# return results:
sum_netout <- getExp.coef.out(object = net_coupler_out_1, metabolite = colnames(met_data_rename))

# get original metabolite names back:
sum_netout <- merge(sum_netout, as.data.frame(met_mapping), by = "Outcome")

# summary statistics:
sum_netout_stat <- mult.stat.surv(sum_netout = sum_netout, adjust_method = "fdr", round_number = 1)


# repeat net.coupler.out with new input and new always set now including all metabolites previously identified as direct effect metabolites:
net_coupler_out_2 <- net.coupler.out(graph_skel = met_SC_skel_rename, dat = met_data_rename_amb1_netout, dat_compl = met_data_rename, exp_dat = Exp_data_DE1_netout, DE = DE, survival_obj = t2d_surv, always_set = always_set_DE1)

sum_netout_2 <- getExp.coef.out(object = net_coupler_out_2, metabolite = colnames(met_data_rename_amb1_netout))

# get original metabolite names back:
sum_netout_2 <- merge(sum_netout_2, as.data.frame(met_mapping), by = "Outcome")

# summary statistics:
sum_netout_stat_2 <- mult.stat.surv(sum_netout = sum_netout_2, adjust_method = "fdr", round_number = 2)


```

