---
title: "Examples of using NetCoupler with different models"
bibliography: ../references.bib
nocite: |
    @Wittenbecher2022, @Wittenbecher2017
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)

make_table <- function(object, caption = NULL) {
    object %>%
        mutate(across(where(is.numeric), round, digits = 3)) %>% 
        knitr::kable(caption = caption)
}
```

This article contains a list of models that you could use in NetCoupler as well
as some published real-world examples of it being used.

There are some caveats to most of these examples, except for when a section
explicitly indicates otherwise:

- Most of the models do not include further adjustment for confounders.
- Most models only show the outcome-side estimation, for brevity. However,
the code for running exposure-side estimation is almost the exact same.

Across the different models, there are some general features, which is described
in more depth in the [Getting Started](netcoupler.html) article.

## Pre-processing

First load up the packages.

```{r}
library(NetCoupler)
library(dplyr)
```

1. Pre-processing by standardizing variables, since large differences in the
values of variables can impact on the results of the algorithm. 

    ```{r}
    standardized_data <- simulated_data %>%
        nc_standardize(starts_with("metabolite"))
    ```

2. Estimating the network structure to identify links between the metabolic variables.
This network needs to be converted into an edge table that has two columns, one 
for the `source_node` and another for the `target_node`.

    ```{r}
    metabolite_network <- simulated_data %>%
        nc_standardize(starts_with("metabolite")) %>% 
        nc_estimate_network(starts_with("metabolite"))
    
    edge_table <- as_edge_tbl(metabolite_network)
    edge_table
    ```

3. If adjusting for confounders in the main models, these also need to be included
when estimating the network. To do this, the metabolic data needs to be
regressed on to the variables that will be standardized.

    ```{r}
    metabolite_network <- simulated_data %>%
        nc_standardize(starts_with("metabolite"),
                       regressed_on = "age") %>%
        nc_estimate_network(starts_with("metabolite"))
```

### Outcome vs exposure side

```{r}
lm_results_outcome <- standardized_data %>%
  nc_estimate_outcome_links(
    edge_tbl = edge_table,
    outcome = "outcome_continuous",
    model_function = lm
  )
```

```{r, echo=FALSE}
make_table(lm_results_outcome$classified_effects)
```

```{r}
lm_results_exposure <- standardized_data %>%
  nc_estimate_exposure_links(
    edge_tbl = edge_table,
    exposure = "exposure",
    model_function = lm
  )
```

## Linear regression

```{r}
lm_results <- standardized_data %>%
  nc_estimate_outcome_links(
    edge_tbl = edge_table,
    outcome = "outcome_continuous",
    model_function = lm
  )
```

```{r, echo=FALSE}
lm_results$classified_effects %>% 
    make_table("Results when using linear models.")
```

## Logistic regression

### Binary Logistic Regression

```{r}
glm_bin_results <- standardized_data %>%
  nc_estimate_outcome_links(
    edge_tbl = edge_table,
    outcome = "case_status",
    model_function = glm,
    model_arg_list = list(family = binomial),
    exponentiate = TRUE,
    implementation = "original"
  )
```

```{r, echo=FALSE}
glm_bin_results$classified_effects %>% 
    make_table("Results when using logistic regression models.")
```

<!-- 
### Ordinal Logistic Regression 

### Nominal Logistic Regression
-->

## Cox Regression

<!-- TODO: Include `cluster()`, `strata()`. -->

```{r}
library(survival)

coxph_results_cluster <- standardized_data %>%
    mutate(
        time_start = age,
        time_end = time_start + outcome_event_time,
        surv_object = Surv(
            time = time_start,
            time2 = time_end,
            event = outcome_binary
        ),
        strata_age = strata(as.integer(age))
    ) %>%
    # summary()
  nc_estimate_outcome_links(
    edge_tbl = edge_table,
    outcome = "surv_object",
    adjustment_vars = "strata_age",
    model_function = survival::coxph,
    # TODO: Test that this works with missingness too.
    model_arg_list = list(cluster = standardized_data$id)
  )

# coxph(Surv() ~ strata(age) + metabolite_1, cluster = id)
```

```{r, echo=FALSE}
coxph_results$classified_effects %>% 
    make_table("Model results and classification for Cox models.")
```

<!-- 
## Poisson regression 

### Negative Binomial Regression
### Zero-inflated regression

## Polynomial Regression

## Ridge Regression

## Lasso Regression

## Quantile Regression

## Elastic Net Regression

## Principal Components Regression

## Support Vector Machines Regression

## Quasi Poisson Regression

## Tobit Regression

## Multivariate Regression

## Random effects models

## Longitudinal data? 

-->

# Real-world uses in published works
